---
title: "Exploring zonation options"
author: "Robyn Seymour-Jones"
date: "26/11/2020"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- '../docs';
  rmarkdown::render(input_file,
  encoding = encoding,
  output_file = file.path(dirname(input_file), out_dir, 'exploring-zonation-options.html'))})
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```




## 1. Introduction
  
I want to define diatom assemblage zones within the record so that it is easier 
to describe and compare with other sequences. This sort of thing is usually done 
by a constrained cluster analysis whereby stratigraphically adjacent samples are 
grouped together based on the similarities of their diatom assemblage. The 
specific cluster analysis chosen is the **Constrained Incremental Sum of Squares 
(CONISS)** method as outlined by Grimm (1987).  
  
  
  
  
### The CONISS algorithm  

First, a matrix of dissimilarities between samples is computed. The CONISS 
algorithm then computes a statistic known as the *sum of squares*\* between each 
pair of adjacent samples (each sample can be considered a cluster of just one 
sample at this stage). The pair with the smallest sum of squares is joined into 
a cluster and then the sum of squares is recalculated for all samples with these
newly joined samples recieving one sum of squares value for their cluster. The
clusters with the smallest sum of squares is joined and the sum of squares
recalulated. This process continues, clustering samples into successively
larger groups (it is therefore an agglomerative technique).

*\*The sum of squares is the squared difference between the value of a taxon in
one sample of a cluster divided by the average value of that taxon across all
samples in that cluster, which then summed for each taxon, which is then
summed for each sample in the cluster.*
  
  
  
  
### Decisions to make

1. I need to decide what to do about one particular outlier sample. If it is not 
good enough for the cluster analysis, I also need to decide if I am going to 
remove it from all of the other analyses completely.

2. Do I perform the cluster analysis with P. ocellata split into their size 
classes as they will be displayed on the stratigraphical diagram or with them 
just grouped as one taxon?

3. Should I perform the cluster analysis on all of the taxa that were counted 
or on only the abundant taxa?

4. If I perform the cluster analysis on only the abundant taxa, should I choose 
those present at 5 % or above in at least one sample as is done by Gordon and 
Birks (1972) or 4 % to keep this analysis consistent with the other results on 
my stratigraphic diagram?

5. If I perform the cluster analysis on only the abundant taxa, should I base 
the percentage abundance calculation on all taxa or only the abundant taxa? The
latter is what is done by Gordon and Birks (1972).



  
### Purpose of this document
  
This document will display the results of various different cluster analyses to 
help me answer the above questions. It also puts them all in one place for easy 
reference at a later date.  
  
  
  
  
### A twist in the tale  
  
To make matters even more complicated, I first performed a cluster analysis on 
the diatom results using the Tilia program (Grimm, 2011). I have described the 
results according to the zones that it produced. To avoid a massive amount of 
repeat work, the final zonation resulting from the cluster analysis needs to be 
as close as possible to the original cluster analysis results I obtained in 
Tilia.




********************************************************************************




## 2. Set up

The data are imported with the manipulate script. I will need the data frames in 
the **taxa** list. I will also need the depths data frame in the **imported** 
list.

```{r message=FALSE}
library(rioja) # chclust()
library(ggdendro) # dendro_data()
library(ggplot2)
library(patchwork)
source("scripts/02-manipulate.R")
```




********************************************************************************




## 3. Data preparation method

The data need to be prepared correctly prior to the cluster analysis. The 
following steps are required.

1. Samples containing no diatoms are removed.

2. The analysis needs to be performed on relative abundance data, not on raw 
counts. These must first be calculated and are done so differently in the 
analyses in this document as I try to figure out if the proportion is out of all 
diatoms counted or only out of abundant taxa. Although I use percentage relative 
abundance for my other analyses and displaying results, the Tilia program 
performs the cluster analysis on relative proportions out of 1.0 so the 
percentage relative abundances are divided by 100 to transform them to unit 
proportions (i.e. out of 1.0). This step is only necessary to keep the result 
the same as the one I obtained from Tilia. If this step is not done, the 
resulting dendrogram looks the same except the x-axis values are just 100 times 
larger.

3. The data are then square-root transformed. This step is necessary as the 
record is dominated by one taxon (P. ocellata, particularly valves ≤20 µm). 
Square-root transforming ensures the less abundant taxa play a more important 
role in the cluster analysis than they would otherwise. My supervisor says this 
is what you must always do and reading around the literature it does seem to be 
standard practice for diatom abundance data.

4. The squared Euclidean distance between samples is then calculated. This 
dissimilarity index is used because it is the one that the Tilia program uses 
and once again I am trying to match the results I previously obtained. Grimm (1987) 
says that the combination of square-root transforming and then calculating the 
squared Euclidean distance has proved "particularly satisfactory" for abundance 
data. As researchers have been using Tilia for years to define zones in 
ecological records, it is therefore also the distance that other researchers 
have been using for the same purpose as me.




********************************************************************************




## 4. Addressing my questions

### Tilia analysis

This reproduces the cluster analysis that I performed in Tilia.

```{r fig.height=9}
# Prepare data
dist_matrix_tilia <- taxa$rel_ab_split_20µm %>%
  mutate(
    depth = imported$depths$depth,
    sample_no = NULL
  ) %>%
  column_to_rownames("depth") %>%
  filter(., rowSums(.) > 0) %>%
  `/`(100) %>%
  sqrt() %>%
  designdist(., method = "A+B-2*J", terms = "quadratic")

# Run cluster analysis
clusters_tilia <- chclust(dist_matrix_tilia, method = "coniss")

# Plot
ddata_tilia <- dendro_data(clusters_tilia, type = "rectangle")

tilia <- ggplot(segment(ddata_tilia)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  coord_flip() +
  scale_x_reverse() +
  labs(x = "Depth (m)",
       y = "Total sum of squares") +
  theme(aspect.ratio = 3) +
  geom_hline(yintercept = 22)

tilia +
  labs(
    title = "Tilia analysis",
    subtitle = "(all taxa,\nP. ocellata 20 µm split,\noutlier still present)"
    )
```




********************************************************************************




### Decision 1: Should I remove the outlier sample?

Several samples in my record were so badly preserved that no diatoms were 
counted. There were a handful of samples adjacent to these where preservation 
was still poor but enough diatoms were present to count around 100 diatom valves. 
Of these samples, all except one had a similar assemblage to those 
stratigraphically adjacent to them so I deemed that although their preservation 
was poor, they were probably representative of their environment. However, this 
one sample in MIS 7d is very different to the others adjacent to it. It has the 
lowest planktonic abundance of the whole record, whilst this might be what was 
actually living in the lake at the time, it is also a possibility that the 
preservation is so poor that it has skewed the assemblage to favour valves that 
preserve better than others. I should probably remove it. But if I do, should I 
remove the other samples where I was only able to count 100 diatoms or is my 
argument that they are similar to their adjacent samples strong enough to keep 
them in?




#### Compare dendrograms

```{r fig.height=9}
# With outlier removed
# ————————————————————

# Prepare data
dist_matrix_no_outlier <- taxa$rel_ab_split_20µm %>%
  mutate(
    depth = imported$depths$depth,
    sample_no = NULL
  ) %>%
  column_to_rownames("depth") %>%
  filter(., rowSums(.) > 0) %>%
  `/`(100) %>%
  sqrt() %>%
  rownames_to_column("depth") %>%
  mutate(depth = as.numeric(depth)) %>%
  slice(., -44) %>%
  column_to_rownames("depth") %>%
  designdist(., method = "A+B-2*J", terms = "quadratic")

# Run cluster analysis
clusters_no_outlier <- chclust(dist_matrix_no_outlier, method = "coniss")

# Create plots
ddata_no_outlier <- dendro_data(clusters_no_outlier, type = "rectangle")
no_outlier <- ggplot(segment(ddata_no_outlier)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  coord_flip() +
  scale_x_reverse() +
  labs(x = "Depth (m)",
       y = "Total sum of squares") +
  theme(aspect.ratio = 3) +
  geom_hline(yintercept = 22)


# Plot dendrograms
# ————————————————
tilia + 
  labs(
    title = "Outlier still present",
    subtitle = "(all taxa,\nP. ocellata 20 µm split,\noutlier still present)"
  ) +
no_outlier +
  labs(
    title = "Outlier removed",
    subtitle = "(all taxa,\nP. ocellata 20 µm split,\noutlier removed)"
  )
```

Removing the outlier has little effect on the rest of the dendrogram. I grouped 
the outlier in with the samples above it in my original description so removing 
the outlier in this situation would be best as it would mean I wouldn't have to 
explain why I have grouped this single sample that should be in its own group in 
with other samples.




#### Compare broken stick

```{r results="hide"}
par(mfrow = c(1, 2))
bstick(clusters_tilia, ng = 25)
bstick(clusters_no_outlier, ng = 25)
```

The broken stick results are annoying as the observed sum of squares fluctuate 
above and below the model at precisely the point at which I want them to go 
below the model for the rest of the plot. Removing the outlier removes the 
seventh point and shifts the remaining points to the left. This makes sense 
because the outlier sample goes into a group on its own on the seventh split 
from the right of the Tilia dendrogram with a total sum of squares of around 
25.5. The broken stick with the outlier suggests there should be 7 or 11 zones 
depending on when you think the observed sum of squares falls below the model. 
The broken stick without the outlier suggests there should be 7 or 10 zones.




#### *Decision: remove outlier*




********************************************************************************




### Decision 2: Should I split *P. ocellata* for the cluster analysis?

Due to the way the Tilia software works, I perfomed the cluster analysis in that 
program on *P. ocellata* split into the 20 µm size classes (≤20 µm, >20 µm and 
unclassifiable). The program required me to input the data with them split as 
the primary function of that program is to plot the stratigraphic diagram. 
However, thinking about this, I don't think it is a good idea to do the cluster 
analysis with them split in this way. Mainly because it creates this artificial 
group of unclassifiable valves (valves that were too dissolved or obscured by 
sediment/other valves to assign a size class too). This artificial group could 
end up influencing where the zone boundaries go and I wouldn't want that at all. 
So here I am just checking that performing the cluster analysis with the valves 
grouped won't affect the zone boundaries too much and cause me to rewrite my 
results section.




#### Compare dendrograms

```{r fig.height=9}
# With no P. ocellata split
# —————————————————————————————

# Prepare data
dist_matrix_no_split <- taxa$rel_ab %>%
  mutate(
    depth = imported$depths$depth,
    sample_no = NULL
  ) %>%
  column_to_rownames("depth") %>%
  filter(., rowSums(.) > 0) %>%
  `/`(100) %>%
  sqrt() %>%
  rownames_to_column("depth") %>%
  mutate(depth = as.numeric(depth)) %>%
  slice(., -44) %>%
  column_to_rownames("depth") %>%
  designdist(., method = "A+B-2*J", terms = "quadratic")

# Run cluster analysis
clusters_no_split <- chclust(dist_matrix_no_split, method = "coniss")

# Create plots
ddata_no_split <- dendro_data(clusters_no_split, type = "rectangle")
no_split <- ggplot(segment(ddata_no_split)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  coord_flip() +
  scale_x_reverse() +
  labs(x = "Depth (m)",
       y = "Total sum of squares") +
  theme(aspect.ratio = 3) +
  geom_hline(yintercept = 21)


# Plot dendrograms
# ————————————————
no_outlier + 
  labs(
    title = "P. ocellata 20 µm split",
    subtitle = "(all taxa,\nP. ocellata 20 µm split,\noutlier removed)"
  ) +
no_split +
  labs(
    title = "No P. ocellata split",
    subtitle = "(all taxa,\nno P. ocellata split,\noutlier removed)"
    )

```

If I move the vertical line at which I am splitting the dendrogram into zones 
(from total sum of square of 22 to 21), I can create the same zones as before. 
The relationships between the zones change slightly but that doesn't matter for 
what I am trying to do.




#### Compare broken stick

```{r}
par(mfrow = c(1, 2))
bstick(clusters_no_outlier, ng = 25)
bstick(clusters_no_split, ng = 25)
```

It does have some effect on the number of zones that the broken stick model 
suggests I should have. It suggests only 6 or 9 zones. Going on this information, 
I probably split my record into too many zones when I did it with Tilia.




#### *Decision: do not split P. ocellata*




********************************************************************************




### Decision 3: Should the cluster analysis be performed on all or abundant taxa?

Gordon and Birks (1972) suggest that the cluster analysis should only be run on 
taxa present at ≥5 % in at least one sample as the low abundance taxa "are of 
little numerical importance in biostratigraphy". Grimm (1987) eliminated those 
present at "<3 % at every level", noting that the "elimitation of rare taxa has 
little effect on the analysis". In the paper about determining the number of 
zones to split a record into, Bennett (1996) also uses only the abundant taxa 
present at >5 % in at least one sample, stating that "Birks & Berglund (1979) and 
Birks (1986) recommended that data sets for numerical zonation should only 
contain those taxa with proportions that exceed 5 %". Bennett (1996) then goes 
on to assess the effect of lowering and increasing the threshold for taxa 
inclusion. Lowering the threshold slight increases the number of zones 
recognised, raising the threshold to 10 % or 20 % changes the location of zone 
markers to positions that suit the remaining taxa. It's interesting to note here 
that using the CONISS method, the analysis in Bennett (1996) identified 6 zones 
with the threshold set at 0-5 %, 1 %, 2 % or 5 %, and identified 5 zones with 
the threshold set at 10 % or 20 % — a difference of only one zone.

Investigating those references provided by Bennett (1996), Birks (1986, pg 757) 
gives the following reason for including on abundance taxa:  

>"Experience with many data sets has shown that pollen types with values less 
than 5 % are of minor *numerical* importance and exert little influence on the 
zonation results. Such rare types have a high relative error associated with 
their pollen counts (Maher, 1972; Faegri and Iversen, 1975) and are thus poorly 
estimated numerically unless very large pollen sums are counted. Such types may, 
of course, be of very considerable *ecological* importance as 'indicator species' 
in the interpretation of the data in terms of past flora, vegetation and 
environment (e.g. Iversenm, 1944, 1954; Birks, 1973, 1976)."

Obviously the work cited so far is quite old so I have looked around to see if 
there has been any progression on this. Birks (2012) states that the basic 
principles "remain largely unchanged" since they were established in Gordon and 
Birks (1972) and Birks and Gordon (1985).

Having read all of this, my conclusion is that I should be running the cluster 
analysis on only abundant taxa. Although running it on all taxa would have 
little impact on the resulting zonation, their inclusion could cause slight 
alterations. These might not be desirable as the data they are based on might 
not be an accurate representation of the population. The abundances of some rare 
taxa could be due to the sampling.




#### Compare dendrograms

```{r fig.height=9}

# On taxa present at ≥4%
# ——————————————————————

# Prepare data
dist_matrix_ab_taxa <- taxa$rel_ab_4 %>%
  mutate(
    depth = imported$depths$depth,
    sample_no = NULL
  ) %>%
  column_to_rownames("depth") %>%
  filter(., rowSums(.) > 0) %>%
  `/`(100) %>%
  sqrt() %>%
  rownames_to_column("depth") %>%
  mutate(depth = as.numeric(depth)) %>%
  slice(., -44) %>%
  column_to_rownames("depth") %>%
  designdist(., method = "A+B-2*J", terms = "quadratic")

# Run cluster analysis
clusters_ab_taxa <- chclust(dist_matrix_ab_taxa, method = "coniss")

# Create plots
ddata_ab_taxa <- dendro_data(clusters_ab_taxa, type = "rectangle")
ab_taxa <- ggplot(segment(ddata_ab_taxa)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  coord_flip() +
  scale_x_reverse() +
  labs(x = "Depth (m)",
       y = "Total sum of squares") +
  theme(aspect.ratio = 3)


# Plot dendrograms
# ————————————————
no_split +
  labs(
    title = "All taxa",
    subtitle = "(all taxa,\nno P. ocellata split,\noutlier removed)"
    ) +
ab_taxa +
  labs(
    title = "Abundant taxa only",
    subtitle = "(taxa ≥4 %,\nno P. ocellata split\noutlier removed)"
    ) +
    geom_hline(yintercept = 12.5)
```

Performing the analysis on abundant taxa only does change the dendrogram quite a 
bit but mostly at the smaller splits. The zones can actually remain the same 
except for zones 2/3 (where some samples in the original zone 2 are moved 
into zone 3) and zones 11/12 (where some samples in the original zone 11 are 
moved into zone 12).




#### Compare broken stick

```{r}
par(mfrow = c(1, 2))
bstick(clusters_no_split, ng = 25)
bstick(clusters_ab_taxa, ng = 25)
```

The broken stick is a bit different here too. It becomes even more difficult to 
interpret with multiple fluctuations of the observed sum of squares above and 
below the model. It identifies between 7 and 11 zones.




#### Reassess zones

```{r fig.height=9}
ab_taxa +
  labs(
    title = "7 zones",
    subtitle = "(taxa ≥4 %,\nno P. ocellata split\noutlier removed)"
    ) +
  geom_hline(yintercept = 15) +
ab_taxa +
  labs(
    title = "11 zones",
    subtitle = "(taxa ≥4 %,\nno P. ocellata split\noutlier removed)"
  ) + 
  geom_hline(yintercept = 13.2)
```

Forming 7 or 11 zones is actually quite difficult as for both there is only a 
very narrow area within which the straight line can be placed in order to split 
into those groups. 11 zones is clearly more similar to the original splits that 
I made on the dendrogram created with Tilia.




#### *Decision: use only abundant taxa*




********************************************************************************




### Decision 4: Should abundant taxa be those at ≥ 4 % or ≥ 5 %?

Gordon and Birks (1972) include taxa present at ≥5 % in at least one sample. As 
I am including taxa present at ≥4 % in at least one sample in my stratigraphic 
diagram, I thought it would be best that everything is consistent. I choose ≥4 % 
as this meant that *Asterionella formosa* was still included on the diagram, 
which could be very useful in interpretations. Displaying taxa present at ≥5 % 
meant this taxon was no longer displayed. The resulting stratigraphic diagram 
also displayed well when taxa present at ≥4 % were kept.

As I mentioned under my introduction paragraph of decision 3, Bennett (1996) 
explores the effect that changing the threshold for taxa inclusion has on the 
zonation and found there to be little effect. 




#### Compare dendrograms

```{r fig.height=9}

# On taxa present at ≥5 %
# ———————————————————————

# Prepare data
taxa$rel_ab_5 <- taxa$rel_ab %>%
  transmute(
    sample_no = sample_no,
    select(., all_of(abundant_taxa(., 5)))
  )

dist_matrix_ab_taxa_5 <- taxa$rel_ab_5 %>%
  mutate(
    depth = imported$depths$depth,
    sample_no = NULL
  ) %>%
  column_to_rownames("depth") %>%
  filter(., rowSums(.) > 0) %>%
  `/`(100) %>%
  sqrt() %>%
  rownames_to_column("depth") %>%
  mutate(depth = as.numeric(depth)) %>%
  slice(., -44) %>%
  column_to_rownames("depth") %>%
  designdist(., method = "A+B-2*J", terms = "quadratic")

# Run cluster analysis
clusters_ab_taxa_5 <- chclust(dist_matrix_ab_taxa_5, method = "coniss")

# Create plots
ddata_ab_taxa_5 <- dendro_data(clusters_ab_taxa_5, type = "rectangle")
ab_taxa_5 <- ggplot(segment(ddata_ab_taxa_5)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  coord_flip() +
  scale_x_reverse() +
  labs(x = "Depth (m)",
       y = "Total sum of squares") +
  theme(aspect.ratio = 3)


# Plot dendrograms
ab_taxa +
  labs(
    title = "≥4 %",
    subtitle = "(taxa ≥4 %,\nno P. ocellata split\noutlier removed)"
  ) +
  geom_hline(yintercept = 12.5) +
ab_taxa_5 +
  labs(
    title = "≥5 %",
    subtitle = "(taxa ≥5 %,\nno P. ocellata split\noutlier removed)"
  ) +
  geom_hline(yintercept = 11.5)
```

The resulting dendrograms are very similar. For the ≥5 % dendrogram, a shift of 
the verical line to the left results in the same zonation as for ≥4 % except for 
one sample in zone 11 of ≥4 %, which finds itself in zone 12 of ≥5 %.




#### Compare broken stick

```{r}
par(mfrow = c(1, 2))
bstick(clusters_ab_taxa, ng = 25)
bstick(clusters_ab_taxa_5, ng = 25)
```

The broken stick is very similar for both. It is difficult to make out if it is 
after the 11th or 12th split that the observed sum of squares drops below the 
modelled sum of squares. Even a closer look shows it is very close:

```{r}
bstick(clusters_ab_taxa_5, ng = 25)
```




#### *Decision: stick with ≥4 %*




********************************************************************************




### Decision 5: Should the % abundance be based on a sum of all taxa or abundant taxa?

My instinct was to calculate percentage abundances of a taxon out of all taxa 
that were counted in a sample. I didn't even think of doing it another way until 
I read that Gordon & Birks (1972) calculated their percentages out of only the 
abundant taxa that they were using in their CONISS analysis. Every single paper 
I have come across has used their method, calculating percentage abundance of 
taxa based only on abundant taxa. I have therefore done the same throughout this 
document. This should have been the first question I addressed but there was no 
need to show the differences that occur in the dendrograms because of this as I 
would never have considered doing it the way I had originally thought because it 
is not the done thing. 

I am struggling with why this would be done but will have a go at guessing why — 
the taxa present at such low abundances are not statistically viable because 
they are potentially only there as a result of sampling chance. It could have 
been a fluke that a valve of a rare taxon fell on the part of the slide I was 
looking at rather than one of the more abundant taxa. Thus, the counts for the 
abundant taxa appear lower than they should have been. Or this might not be the 
case and the occurence of that rare taxon valve is representative of the original 
population. Either way there is an element of doubt due to such low counts 
(sometimes only one or two valves) that it statistically more robust to remove 
them.




#### *Decision: base percentage calculations on sum of only abundant taxa*





********************************************************************************




## 5. Conclusion

After addressing all of my questions, this is the resulting CONISS analysis and 
zonation scheme that I have decided upon.

```{r fig.height=9}
ab_taxa +
  labs(
    title = "Final dendrogram",
    subtitle = "(taxa ≥4 %,\nno P. ocellata split\noutlier removed)"
  ) +
  geom_hline(yintercept = 13.1)
```

```{r fig.height=9}
bstick(clusters_ab_taxa, ng = 25)
```





********************************************************************************




## References

Bennett, K. D. (1996) Determination of the number of zones in a 
biostratigraphical sequence. *New Phytologist*, 132, 155–170.

Birks, H. J. B. (1986) Numerical zonation, comparison and correlation of 
Quaternary pollen-stratigraphical data. In Berglund, B. E. (ed) 
*Handbook of Palaeoecology and Palaeohydrology*. Chichester: John Wiley & Sons.

Birks, H. J. B. (2012) Analysis of stratigraphical data. In Birks, H. J. B., 
Lotter, A. F., Juggins, S. & Smol, J. P. (eds) *Tracking Environmental Change* 
*using Lake Sediments. Volume 5: Data Handling and Numerical Techniques*. 
Dordrecht: Springer.

Birks, H. J. B. & Berglund, B. E. (1979) Holocene pollen stratigraphy of southern 
Sweden: a reappraisal using numerical methods. *Boreas*, 8, 257–279.

Birks, H. J. B. & Gordon, A. D. (1985) *Numerical methods in Quaternary pollen analysis*. 
London: Academic Press.

Gordon, A. D. & Birks, H. J. B. (1972) Numerical methods in Quaternary 
palaeoecology. I. Zonation of pollen diagrams. *New Phytologist*, 71, 961–979.  

Grimm, E. C. (1987) CONISS: A Fortran 77 program for stratigraphically 
constrained cluster analysis by the method of incremental sum of squares. 
*Computers & Geosciences*, 13, 13–35.

Grimm, E. C. (2011) *Tilia version 1.7.16* [Software]. Illinois State Museum, Springfield.
