---
title: "Exploring ordination options"
author: "Robyn Seymour-Jones"
date: "15/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Introduction
I want to perform an ordination on my diatom results to pick out the main 
variation in the diatom assemblage throughout my record and to see if I can find 
out what might be driving this variation. I don't have any environmental data so 
I won't be trying to fit any known variables to the ordination. Instead, I will 
use knowledge of the environmental preferences of the taxa along the gradients 
to interpret what environmental variables the gradients could represent. 
Therefore, I will only be looking into unconstrained ordination methods. There 
are several options to decide upon when running the ordination:

* Do the data need to be transformed?


* What type of ordination should be performed? 
I have already decided I need to use unconstrained ordination techniques. But I
also need to decide on the type of ordination to use based on the distributions 
of the species data. 


* Should the raw diatom counts be transformed? If so, which transformation 
should be used?

* Should the diatom data be scaled to unit variance (i.e. standardised)?



## Set up

First the requirements are loaded. The manipulate script loads the raw data into 
a list of dataframes (imported) and then manipulates the data into various other 
lists of dataframes.

```{r message=FALSE}
source("scripts/02-manipulate.R")
library(ggplot2)
```

This is the structure of the data and functions that area loaded by running the 
manipulate script.

```{r}
ls.str()
```

The ordination data will go in its own list so this is created next. 

```{r}
ordination <- list()
```

The dataframes have sample_no as the first variable and still contain samples 
within which no diatoms were counted. These must be removed for ordination. A
function is created so that the original dataframes can remain in a consistent
format and so that I always know that any dataframe I come to use will have
sample_no as the first variable. The function removes samples with no diatoms 
and moves sample_no to row names so that this information is preserved after 
ordination.

```{r}
ord_data <- function(df) {
  df %>%
    filter(rowSums(.[, -1]) > 0) %>%
    column_to_rownames("sample_no")
}
```

## Ordination of all taxa vs. abundant taxa

## Ordination of different transfromations

## Plotting difference scalings






## Ordination on abundant taxa only
