---
title: "Exploring ordination options"
author: "Robyn Seymour-Jones"
date: "09/12/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: lumen
knit: (function(input_file, encoding) {
  out_dir <- '../docs';
  rmarkdown::render(input_file,
  encoding = encoding,
  output_file = file.path(dirname(input_file), out_dir, 'exploring-ordination-options.html'))})
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## 1. Introduction
I want to perform an ordination on my diatom results to pick out the main 
variation in the diatom assemblage throughout my record and to see if I can find 
out what might be driving this variation. I don't have any environmental data so 
I won't be trying to fit any known variables to the ordination. Instead, I will 
use knowledge of the environmental preferences of the taxa along the gradients 
to interpret what environmental variables the gradients could represent. 
Therefore, I will only be looking into unconstrained ordination methods.  
  
There are several options to decide upon when running the ordination:

1. Should the raw diatom counts be transformed? If so, which transformation 
should be used?

2. Should the diatom data be scaled to unit variance (i.e. standardised)?

3. Should the ordination be performed on all taxa or only abundant taxa (e.g. 
taxa present at ≥4% in at least one sample)? If performing on only abundant taxa, 
should the relative abundance be calculated from a sum of all taxa or only those 
abundant taxa?

4. What type of ordination should be performed? I have already decided I need to 
use unconstrained ordination techniques. But I also need to decide on the type 
of ordination to use based on the distributions of the species data.

********************************************************************************




## 2. Set up

The data are imported with the 
[manipulate script](https://github.com/robynfsj/phd-ioannina-long-record/blob/master/scripts/02-manipulate.R). 
I will need the data frames in the **taxa** list. I will also need the depths 
data frame in the **imported** list.

```{r message=FALSE}
library(ggplot2)
library(patchwork)
source("scripts/02-manipulate.R")
```




********************************************************************************




## 3. Data preparation method

The dataframes have sample_no as the first variable and still contain samples 
within which no diatoms were counted (as well as an outlier where diatoms were 
present, but only 100 were counted). These must be removed for ordination. A
function is created so that the original dataframes can remain in a consistent
format and so that I always know that any dataframe I come to use will have
sample_no as the first variable. The function removes samples with no diatoms 
and moves sample_no to row names so that this information is preserved after 
ordination.

```{r}
ord_data <- function(df) {
  df %>%
    filter(sample_no != "189" & rowSums(.[, -1]) > 0) %>%
    column_to_rownames("sample_no")
}
```




********************************************************************************




## 4. Addressing my questions

### Decision 1: Should the raw diatom counts be transformed? And if so, which transformation should be used?

[Legendre and Borcard (2018:1821)](#references) state that "[e]cologists have 
long known that untransformed (site × species) community composition data should 
not be analysed using the Euclidean distance, or through methods of multivariate 
analysis that preserve the Euclidean distance, such as principal component 
analysis and redundancy analysis, where calculations are carried out in 
Euclidean space". Therefore, if I am to perform a principal component analysis 
(PCA), it is clear that I need to transform the data prior to the analysis. 

Transforming the data reduces the skewness of a species distribution and helps 
to address the double zero problem (the unwanted situation where the absence of 
a species from two sites is interpreted as an indication of similarity between 
the sites). Through doing this, it makes the data suitable for analysis using 
linear methods (such as PCA) within Euclidean space.

I don't think I am that interested in placing higher weightings on rare taxa as 
I am looking to find the main variation within the data. If the record is 
dominated by one taxon (as I know it is), I want the results of the ordination 
to show this. However, I don't want the absence of a species from two samples to 
indicate similarity.




#### Transformation: relative abundances

I have already transformed the data into percentage relative abundances as part 
of the [manipulate script](https://github.com/robynfsj/phd-ioannina-long-record/blob/master/scripts/02-manipulate.R). 
These are the data frames in the taxa list with the rel_ab prefix. Note that I 
perform these analyses on abundant taxa only with percentages calculated based 
on a sum of all taxa counted in a sample. This is so that the analyses produce 
cleaner plots with fewer taxa displayed. I will address whether this is the 
right thing to do in later sections of this document.

##### DCA

```{r}
untransformed_dca <- decorana(ord_data(taxa$rel_ab_4))
untransformed_dca
```

DCA1 axis length is <2.5 so linear methods such as PCA are appropriate 
([Legendre & Birks, 2012](#references)). 

##### PCA

```{r}
untransformed_pca <- rda(ord_data(taxa$rel_ab_4))
untransformed_pca
```

```{r collapse=TRUE}
# Proportion of variance explained by PC1 and PC2
sum((as.vector(untransformed_pca$CA$eig) / sum(untransformed_pca$CA$eig))[1:2]) * 100

# Proportion of variance explained by PC1
sum((as.vector(untransformed_pca$CA$eig) / sum(untransformed_pca$CA$eig))[1]) * 100

# Proportion of variance explained by PC2
sum((as.vector(untransformed_pca$CA$eig) / sum(untransformed_pca$CA$eig))[2]) * 100

```




##### Plot of PC1 by core depth

```{r message=FALSE}
untransformed_pc1_scrs <- data.frame(scores(untransformed_pca, 
                                            choices = 1, # select only PC1
                                            display = "sites",
                                            scaling = 0)) %>%
  mutate(sample_no = as.numeric(row.names(.))) %>%
  right_join(imported$depths)

plot_untransformed_pc1 <- ggplot(untransformed_pc1_scrs, aes(x = depth, y = PC1)) +
  geom_line() +
  coord_flip() +
  scale_x_reverse(expand = c (0, 0), breaks = seq(130, 290, 10)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Depth (m)") +
  theme_classic(8) +
  theme(aspect.ratio = 3) 

plot_untransformed_pc1 
```




##### Quick ordination biplots

```{r}
biplot(untransformed_pca, scaling = "none")
```

Scale the species in order to display the distribution of the sites more clearly.

```{r}
biplot(untransformed_pca, scaling = "species")
```




##### Evaluation

According to [Legendre and Gallagher (2001)](#references), a species profile 
transformation is not the best transformation to use as it still considers 
double zeros to be an indicator of similarity.

PC1 and PC2 explain a large proportion of the variance in the PCA (81%). In 
particular, PC1 explains 65% of the variance, which is good as I want to plot 
PC1 alongside my stratigraphic diagram.

The sites are spread out oddly, forming a sort of wedge shape pointing towards 
the left of the plot and two trailing tail shapes on the right side of the plot, 
which I think might be the horseshoe effect. I don't really understand the 
horseshoe effect except that it isn't desirable. I don't know if the horseshoe 
shape is always vertical or if it can appear horizontal as I think it does here.

There are two species that stand out as driving the variation across the samples. 
I have plotted the quick biplots as points above so they are easy to read but 
plotting with text shows these species are *P. ocellata* and *P. minuscula*, 
which is what I expected as *P. ocellata* dominates my record.


```{r}
biplot(untransformed_pca, scaling = "sites", type = "text")
```




********************************************************************************




#### Transformation: Hellinger (square root of relative abundances)

Square root transformation "lowers the importance of the most abundant species, 
which may grow exponentially when they encounter favourable conditions" 
([Legendre & Birks, 2012:216](#references)).




##### Square-root transform data

```{r}
sqrt_rel_ab_4 <- ord_data(taxa$rel_ab_4) %>%
  sqrt()
```




##### DCA

```{r}
sqrt_dca <- decorana(sqrt_rel_ab_4)
sqrt_dca
```
DCA1 axis length is <2.5 so linear methods such as PCA are appropriate 
([Legendre & Birks, 2012](#references)).




##### PCA

```{r}
sqrt_pca <- rda(sqrt_rel_ab_4)
sqrt_pca
```

```{r collapse=TRUE}
# Proportion of variance explained by PC1 and PC2
sum((as.vector(sqrt_pca$CA$eig) / sum(sqrt_pca$CA$eig))[1:2]) * 100

# Proportion of variance explained by PC1
sum((as.vector(sqrt_pca$CA$eig) / sum(sqrt_pca$CA$eig))[1]) * 100

# Proportion of variance explained by PC2
sum((as.vector(sqrt_pca$CA$eig) / sum(sqrt_pca$CA$eig))[2]) * 100
```




##### Plot of PC1 by core depth

```{r message=FALSE}
sqrt_pc1_scrs <- data.frame(scores(sqrt_pca, 
                                   choices = 1, # select only PC1
                                   display = "sites",
                                   scaling = 0)) %>%
  mutate(sample_no = as.numeric(row.names(.))) %>%
  right_join(imported$depths)

plot_sqrt_pc1 <- ggplot(sqrt_pc1_scrs, aes(x = depth, y = PC1)) +
  geom_line() +
  coord_flip() +
  scale_x_reverse(expand = c (0, 0), breaks = seq(130, 290, 10)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Depth (m)") +
  theme_classic(8) +
  theme(aspect.ratio = 3) 

plot_sqrt_pc1
```




##### Quick ordination biplot

```{r}
biplot(sqrt_pca, scaling = "none")
```

Scale the species in order to display the distribution of the sites more clearly.

```{r}
biplot(sqrt_pca, scaling = "species")
```




##### Evaluation

[Legendre and Gallagher (2001)](#references) recommend this transformation 
(or the chord transformation) for species abundance data where it is not desired 
to give large weightings to rare species. According to what I have read, this 
should be the transformation most suitable for my data. 

PC1 only explains 26% of the variance, which is not ideal when I want to plot 
this alongside my stratigraphic diagram as it doesn't explain much.

The biplot looks ok with sites spread out nicely, although there is a slight 
horseshoe effect. Two species drive more variation than the rest. Again these 
are *P. ocellata* and *P. minuscula*:

```{r}
biplot(sqrt_pca, scaling = "sites", type = "text")
```




##### Comparison of transformations 

```{r}
plot_untransformed_pc1 +
  labs(title = "PC1 of percentage\nabundance PCA") +
plot_sqrt_pc1 +
  labs(title = "PC1 of square-root transformed\npercentage abundance PCA")
```


```{r}
par(mfrow = c(1, 2))

biplot(untransformed_pca, 
       scaling = "species", 
       main = "PCA on percentage abundance",
       cex.main = 0.7)

biplot(sqrt_pca, 
       scaling = "species", 
       main = "PCA on square-root transformed percentage abundance",
       cex.main = 0.7)
```


The biplot for the square-root transformation looks much nicer than the biplot 
for the percentage abundance data. The samples are more spread out and the 
species arrows are more similar in length to each other, rather than there being 
two arrows that were much longer than the others. The latter demonstrates the 
impact of square-root transforming the percentage abundances, giving less weight 
to very abundant taxa.





********************************************************************************




### Decision 2: Should the diatom data be scaled to unit variance (i.e. standardised)?

Standardisation is a special form of transformation where the variables are 
transformed to have a mean of 0 and a variance (or sometimes a standard 
deviation) of 1. This puts different variables on the same scale. When working 
with physical, chemical or geological variables, it is standard practice to 
standardise these variables prior to an ordination as they are measured in 
different units. If this step is not done then a variable that is measured in 
larger numbers is weighted more heaviliy in the ordination than a variable 
measured in smaller numbers. For example, lake temperature measured in degrees 
celcius might fall within the range of 5 to 15 degrees but depth could be several 
hundreds of meters. If we want to give the same weighting to these variables so 
they equally contribute to the ordination then they would need to be standardised. 
However, when working with species variables, these are all clearly measured in 
the same unit so there is no need to standardise. But, if the raw counts are 
transformed into relative proportions (and the ordination is run on all taxa, 
not only on the abundant ones), then this is effectively a standardisation.

My question of standardisation here has developed slightly and is only relevant 
if I decide to run the ordination on only abundant taxa. In this situation, do I 
perform the ordination on percentages of the abundant taxa based on a sum of all 
taxa or the percentages based on a sum of only those abundant taxa? This is a 
similar situation to when I was trying to decide on how to run my cluster 
analysis (see decision 5 of [Exploring zonation options](https://robynfsj.github.io/phd-ioannina-long-record/exploring-zonation-options.html)). 

```{r}
# Prepare data for analysis
abundant_taxa <- taxa$counts %>%
  # isolate taxa present at ≥4 %
  transmute(
    sample_no = sample_no,
    select(., all_of(abundant_taxa(taxa$rel_ab, 4)))
  ) %>%
  # convert to proportion based on only these abundant taxa
  mutate(
    sample_no = sample_no,
    decostand(select(., -sample_no), 
              method = "total",
              na.rm = "TRUE")
  )
```

I will look at this for both percentage abundance data and for square-root 
transformed percentage abundance data as I still don't know which I will be 
using yet.




********************************************************************************




#### Transformation: relative abundances

```{r}
ab_t <- ord_data(abundant_taxa)
```




##### DCA

```{r}
ab_t_dca <- decorana(ab_t)
ab_t_dca
```

DCA axis 1 length is <2.5 so linear methods are appropriate. 

```{r}
ab_t_pca <- rda(ab_t)
ab_t_pca
```




##### PCA

```{r collapse=TRUE}
# Proportion of variance explained by PC1 and PC2
sum((as.vector(ab_t_pca$CA$eig) / sum(ab_t_pca$CA$eig))[1:2]) * 100

# Proportion of variance explained by PC1
sum((as.vector(ab_t_pca$CA$eig) / sum(ab_t_pca$CA$eig))[1]) * 100

# Proportion of variance explained by PC2
sum((as.vector(ab_t_pca$CA$eig) / sum(ab_t_pca$CA$eig))[2]) * 100

```

The proportion expained by PC1 and PC2 is 77% so less than the 80% explained by 
the first two axes when the percentages were calculated from a sum of all taxa. 
Proportion explained by PC1 for that analysis was 64%, higher than the 59% 
explained here.




##### PC1 axis scores

```{r message=FALSE}
ab_t_pc1_scrs <- data.frame(scores(ab_t_pca, 
                              choices = 1, # select only PC1
                              display = "sites",
                              scaling = 0)) %>%
  mutate(sample_no = as.numeric(row.names(.))) %>%
  right_join(imported$depths)
```

```{r}
plot_ab_t_pc1 <- ggplot(ab_t_pc1_scrs, aes(x = depth, y = PC1)) +
  geom_line() +
  coord_flip() +
  scale_x_reverse(expand = c (0, 0), breaks = seq(130, 290, 10)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Depth (m)") +
  theme_classic(8) +
  theme(aspect.ratio = 3) 

plot_untransformed_pc1 +
  labs(title = "Percentage abundance\n(% based on all taxa)") +
plot_ab_t_pc1 +
  labs(title = "Percentage abundance\n(% based on abundant taxa)")
```

The PC1 axis scores are very similar. There are a few differences that can be 
seen if you look very carefully. These are most noticable at around 190 m to 
210 m depth, although there are some other slight variations in other places.




##### Biplots

```{r}
par(mfrow = c(1, 2))

biplot(untransformed_pca, 
       scaling = "none", 
       main = "Percentage abundance\n(% based on all taxa)",
       cex.main = 0.7)

biplot(ab_t_pca, 
       scaling = "none", 
       main = "Percentage abundance\n(% based on abundant taxa)",
       cex.main = 0.7)
```

If anything, there is a stronger arch effect for the percentages based on a sum 
of only the abundant taxa.




********************************************************************************




#### Transformation: Hellinger (square root of relative abundances)

```{r}
ab_t_sqrt <- ord_data(abundant_taxa) %>%
  sqrt()
```




##### DCA

```{r}
ab_t_sqrt_dca <- decorana(ab_t_sqrt)
ab_t_sqrt_dca
```

DCA1 <2.5 so linear methods are appropriate.




##### PCA

```{r}
ab_t_sqrt_pca <- rda(ab_t_sqrt)
ab_t_sqrt_pca
```

```{r collapse=TRUE}
# Proportion of variance explained by PC1 and PC2
sum((as.vector(ab_t_sqrt_pca$CA$eig) / sum(ab_t_sqrt_pca$CA$eig))[1:2]) * 100

# Proportion of variance explained by PC1
sum((as.vector(ab_t_sqrt_pca$CA$eig) / sum(ab_t_sqrt_pca$CA$eig))[1]) * 100

# Proportion of variance explained by PC2
sum((as.vector(ab_t_sqrt_pca$CA$eig) / sum(ab_t_sqrt_pca$CA$eig))[2]) * 100

```

The percentages are very similar as for when the percentage abundance was based 
on a sum of all taxa.




##### PC1 axis scores

```{r message=FALSE}
ab_t_sqrt_pc1_scrs <- data.frame(scores(ab_t_sqrt_pca, 
                              choices = 1, # select only PC1
                              display = "sites",
                              scaling = 0)) %>%
  mutate(sample_no = as.numeric(row.names(.))) %>%
  right_join(imported$depths)
```

```{r}
plot_ab_t_sqrt_pc1 <- ggplot(ab_t_sqrt_pc1_scrs, aes(x = depth, y = PC1)) +
  geom_line() +
  coord_flip() +
  scale_x_reverse(expand = c (0, 0), breaks = seq(130, 290, 10)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Depth (m)") +
  theme_classic(8) +
  theme(aspect.ratio = 3) 

plot_sqrt_pc1 +
  labs(title = "sqrt, % based on all taxa") +
plot_ab_t_sqrt_pc1 +
  labs(title = "sqrt, % based on abundant taxa")
```

There isn't much difference between the two, although it has flipped the results. 
If I reverse the PC1 axis values for one of them then it is possible to see just 
how similar they are:

```{r}
plot_sqrt_pc1_rev <- ggplot(sqrt_pc1_scrs, aes(x = depth, y = PC1)) +
  geom_line() +
  coord_flip() +
  scale_x_reverse(expand = c (0, 0), breaks = seq(130, 290, 10)) +
  scale_y_reverse(expand = c(0, 0)) +
  labs(x = "Depth (m)") +
  theme_classic(8) +
  theme(aspect.ratio = 3) 


plot_sqrt_pc1_rev +
  labs(title = "sqrt, % based on all taxa") +
plot_ab_t_sqrt_pc1 +
  labs(title = "sqrt, % based on abundant taxa")
```

The are almost identical. It is possible to see a very slight difference in the 
values at around 220 m depth where the values do not quite reach the y-axis on 
the right-hand plot.




##### Biplots

```{r}
par(mfrow = c(1, 2))

biplot(sqrt_pca, 
       scaling = "none", 
       main = "sqrt, % of all taxa",
       cex.main = 0.7)

biplot(ab_t_sqrt_pca, 
       scaling = "none", 
       main = "sqrt, % of ab. taxa",
       cex.main = 0.7)
```

Plots are very similar but rotated 180 degrees.




********************************************************************************



## References

Legendre, P & Birks, H. J. B. (2012) From classical to canonical ordination. In 
Birks, H. J. B., Lotter, A. F., Juggins, S. & Smol, J. P. (eds) *Tracking *
*Environmental Change using Lake Sediments. Volume 5: Data Handling and *
*Numerical Techniques*. Dordrecht: Springer.
https://doi.org/10.1007/978-94-007-2745-8_8

Legendre, P. & Borcard, D. (2018) Box–Cox-chord transformations for community 
composition data prior to beta diversity analysis. *Ecography*, 41, 1820–1824. 
https://doi.org/10.1111/ecog.03498

