---
title: "Plotting dendrograms"
author: "Robyn Seymour-Jones"
date: "04/12/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: lumen
knit: (function(input_file, encoding) {
  out_dir <- "../docs";
  rmarkdown::render(input_file,
  encoding = encoding,
  output_file = file.path(dirname(input_file), out_dir, "plotting-dendrograms.html"))})
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Introduction

It is proving difficult to plot a dendrogram for my diatom record as it contains 
missing samples (some completely missing because of a misplaced section of the 
sediment core and some where there were no diatoms present). The document 
explores the various options for plotting dendrograms in R.

I want to add the dendrogram to the right of my stratigraphic plot, which I am 
plotting with ggplot and the help of the tidypaleo package. Therefore, I ideally 
want to plot the dendrogram with ggplot too.




## Load requirements

```{r message=FALSE}
library(rioja)
library(ggplot2)
library(ggdendro)
source("scripts/borders-for-ggplot2.R")
source("scripts/02-manipulate.R")
```




## Prepare data and run cluster analysis

```{r}

# Prepare data for analysis
coniss_data <- taxa$counts %>%
  # isolate taxa present at ≥4 %
  transmute(
    sample_no = sample_no,
    select(., all_of(abundant_taxa(taxa$rel_ab, 4)))
  ) %>%
  # convert to proportion based on only these abundant taxa
  mutate(
    sample_no = sample_no,
    decostand(select(., -sample_no), 
              method = "total",
              na.rm = "TRUE")
  ) %>%
  # replace sample numbers with depths
  mutate(
    depth = imported$depths$depth,
    sample_no = NULL
  ) %>%
  # remove samples with no diatoms
  column_to_rownames("depth") %>%
  filter(., rowSums(.) > 0) %>%
  # square-root transform
  sqrt() %>%
  # remove outlier
  rownames_to_column("depth") %>%
  mutate(depth = as.numeric(depth)) %>%
  slice(., -44) %>%
  column_to_rownames("depth")

# Calculate distance matrix
dist_matrix <- designdist(coniss_data, method = "A+B-2*J", terms = "quadratic")

# Run cluster analysis
coniss_results <- chclust(dist_matrix, method = "coniss")
```




## Plotting with ```plot.chclust()```
I have successfully managed to plot the dendrogram I am after through the 
```plot.chclust()``` method from the rioja package. Although this is what I am 
after, it is not ideal as my other plots are created with ggplot and it makes it 
difficult to combine them in to one plot.

```{r fig.height=9, fig.width=3, fig.align="center"}

# Isolate depths
coniss_depths <- coniss_data %>%
  rownames_to_column("depth") %>%
  mutate(depth = as.numeric(depth)) %>%
  select(., "depth")

# Plot
plot(
  coniss_results, 
  hang = -1, # make leaves extend to 0 and labels hang down from there
  horiz = TRUE, # display plot vertically (depths now "x-axis")
  x.rev = TRUE, # reverse depths
  xvar = coniss_depths$depth, # plot samples by depth rather than just sequential order
  labels = FALSE, # remove leaf labels
  cex.axis = 0.6,
  tcl = -0.1
)
```

There are a few other problems with this plot. I can't plot the axis label (it 
should be "Total sum of squares"). Also, the leaves extend to the left of the 0 
but I would ideally want them to stop at 0. It doesn't seem to do this in the 
examples I have seen elsewhere.




## Plotting with the ggdendro package

This is the first way I plotted my dendrograms but I just couldn't figure out 
how to plot the leaves/segments by depth rather than just in sequential order. I 
think it would require some sort of alteration to the ```dendro_data()``` 
function in order to extract values for the segments that are aligned with the 
depths rather than just in order. This is something I just don't have the time 
to try to explore.

```{r fig.height=7}

# Extract dendrogram data 
ddata <- dendro_data(coniss_results, type = "rectangle")

# Plot
ggplot(segment(ddata)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  coord_flip() +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_reverse(breaks = NULL,
                  labels = NULL) +
  labs(x = "",
       y = "Total sum of squares") +
  theme_bw(8) +
  theme(aspect.ratio = 4,
        legend.position = "none",
        panel.grid = element_blank(),
        panel.border = theme_border("bottom"),
        panel.background = element_rect(fill = "transparent"),
        plot.background = element_rect(fill = "transparent"))
```

This is exactly how I want the dendrogram to display, except that I want it with 
the gaps that are present in the version created with ```plot.chclust()```.




## Solution from a hero on Stack Overflow

I just could not figure out to plot the dendrogram leaves by depth so I posted 
my first ever question on Stack Overflow. I didn't think I would get a reply 
but I got one with 10 minutes! I had been reluctant to ask a question on there 
as I knew I had to write out the question really clearly and provide example 
data and examples of what I want to achieve. I just thought it would take too 
long and migtht not get a reply. I can't believe how easy it was and how helpful. 
Here is the soloution. It is only three lines of code that was causing me all 
this stress!

```{r fig.height=7}

# Map depths to x values
new_x <- approxfun(ddata$labels$x, as.numeric(as.character(ddata$labels$label))) 
ddata$segments$x <- new_x(ddata$segments$x)
ddata$segments$xend <- new_x(ddata$segments$xend)

# Plot (same as previous)
ggplot(segment(ddata)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  coord_flip() +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_reverse(breaks = NULL,
                  labels = NULL) +
  labs(x = "",
       y = "Total sum of squares") +
  theme_bw(8) +
  theme(aspect.ratio = 4,
        legend.position = "none",
        panel.grid = element_blank(),
        panel.border = theme_border("bottom"),
        panel.background = element_rect(fill = "transparent"),
        plot.background = element_rect(fill = "transparent"))
```





